---
title: "The PlantTracker 'trackSpp()' function"
author: "Alice Stears"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Suggested_PlantTracker_Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  rmarkdown.html_vignette.check_title = FALSE
)
```

```{r setup, include = FALSE}
library(PlantTracker)
library(ggplot2)
library(sf)
```

## Introduction
This vignette dives into the `trackSpp` function, the main 'workhorse' function
in the PlantTracker R package. `trackSpp` transforms a dataset of annual 
maps of plant occurence into a demographic dataset. To accomplish this, the 
function compares maps accross sampling years and assigns unique identifiers to 
plants that overlap from year to year. These unique identifiers are then used to 
assign survival, growth, recruit status, and age to each individual plant in 
each year. 

This process is complex and requires certain assumptions, so the following pages 
will explain and illustrate the logic of each of these steps. We recommend you 
read through this vignette before using trackSpp() in order to fully understand 
the assumptions inherent to the function, and to make sure that you are 
adjusting the user-specificed arguments correctly. 

## *1.* Prepare data
