---
title: "Using the `plantTracker` `trackSpp()` function"
date: 17 February 2022
output:
  minidown::mini_document:
    framework: sakura
    theme: default
    toc: true
    toc_float: true
    toc_highlight: true
vignette: >
  %\VignetteIndexEntry{Using_the_plantTracker_trackSpp_function}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  rmarkdown.html_vignette.check_title = FALSE
)
```

```{r setup, include = FALSE}
library(plantTracker)
library(ggplot2)
library(sf)
# library(devtools)
# load_all()
```

# Introduction

This vignette gives detailed information about the `trackSpp()` function, the 
main "workhorse" function in the `plantTracker` R package. `trackSpp()` transforms 
a data set of annual maps of plant occurrence into a demographic data set. To
accomplish this, the function compares maps across sampling years and
assigns unique identifiers ("trackIDs") to plants that overlap from year
to year. Plants with the same trackID are assumed to be the same
individual. These trackIDs are then used to assign survival, growth,
recruit status, and age to each individual plant in each year.

This process is complex and requires certain assumptions, so the
following pages will explain and illustrate the logic of each of these
steps. We recommend you read through this vignette before using
`trackSpp()` in order to fully understand the assumptions inherent to the
function, and to make sure that you are adjusting the user-specified
arguments correctly.

# *1* Input data

The required inputs to the `trackSpp()` function are explained in detail
in [Suggested `plantTracker`
Workflow](Suggested_plantTracker_Workflow.html), Parts 1.1, 1.2, and 2, as well 
as the "help" file for this function (which you can access by typing `?trackSpp`
in the R console).  However, I'll include a short description of the 
arguments here: 
                     
+--------------+--------------------------------------+---------+--------+
|`trackSpp()`  | description                          |required?|default?|
|argument      |                                      |         |        |
+==============+======================================+=========+========+
| dat          | An sf data frame in which each row   |  Yes    |  N/A   |
|              | has spatial data for an individual   |         |        |
|              | observation in one year.             |         |        |
+--------------+--------------------------------------+---------+--------+
| inv          | A named list in which the name of    |  Yes    |  N/A   |
|              | each element of the list is a quadrat|         |        |
|              | name in `dat`, and the contents of   |         |        |
|              | that list element is  a numeric      |         |        |
|              | vector of all of the years in which  |         |        |
|              | that quadrat was actually sampled    |         |        |
|              | (not just the years that have data   |         |        |
|              |in `dat`!)                            |         |        |
+--------------+--------------------------------------+---------+--------+
| dorm         | A single value greater than or equal |  Yes    |  N/A   |
|              | to 0 indicating  the number of years |         |        |
|              | these species are allowed to go      |         |        |
|              | dormant. **OR** a data frame with a  |         |        |
|              | row for each species in `dat`,       |         |        |
|              | species names in the "Species" column|         |        |
|              | and a dormancy value in the          |         |        |
|              | "dorm" column.                       |         |        |
+--------------+--------------------------------------+---------+--------+
| buff         | A single value greater than or equal |  Yes    |  N/A   |
|              | equal to zero, indicating how  far a |         |        |
|              | far a polygon can move from year `i` |         |        |
|              | to year `i+1` and still be considered|         |        |
|              | the same individual. **OR** a        |         |        |
|              | data frame with a row for each       |         |        |
|              | species present in `dat`, species    |         |        |
|              | names in the "Species"  column, and a|         |        |
|              | `buff` value in the "buff" column.   |         |        |
+--------------+--------------------------------------+---------+--------+
| clonal       | A logical value (TRUE or FALSE)      |  Yes    |  N/A   |
|              | indicating whether a species is      |         |        |
|              | allowed to be clonal or not. **OR**  |         |        |
|              | a data frame with a row for each     |         |        |
|              | species in `dat`, species names in   |         |        |
|              | the "Species" column, and a `clonal` |         |        |
|              | value in  the "clonal" column.       |         |        |
+--------------+--------------------------------------+---------+--------+
| buffGenet    | A single value greater than or equal | only if |  N/A   |
|              | to zero indicating how close polygons| `clonal`|        |
|              | must be to one another in the same   |`= TRUE` |        |
|              | year to be grouped as a genet.       |         |        |
|              | **OR** a data frame with a row       |         |        |
|              | for each species in `dat`, species   |         |        |
|              | names in the "Species" column, and a |         |        |
|              | `buffGenet` value in the "buffGenet" |         |        |
|              | column.                              |         |        |
+--------------+--------------------------------------+---------+--------+
| species/     | Five separate arguments, each a      |  No     |"Species"|
| site/        | character string that indicates the  |         | /"Site"|
| quad/        | name of the column in `dat` that     |         | /"Quad"|
| year/        | contains data for each of these      |         | /"Year"|
| geometry     | required data types. No value is     |         | /"geometry"|
|              | required if the column name is the   |         |        |
|              | same as the default. If only one     |         |        |
|              |column names is different than the    |         |        |
|              |default, then you only need to supply |         |        |
|              |a value for that argument.            |         |        |
+--------------+--------------------------------------+---------+--------+
| aggByGenet   | A logical argument (TRUE or FALSE)   |  No     | TRUE   |
|              | that determines whether the output   |         |        |
|              | will be aggregated by genet.         |         |        |
+--------------+--------------------------------------+---------+--------+
| printMessages| A logical argument (TRUE or FALSE)   |  No     | TRUE   |
|              |that determines if the function       |         |        |
|              |returns informative messages.         |         |        |
+--------------+--------------------------------------+---------+--------+
| flagSuspects | A logical argument (TRUE or FALSE)   |  No     | FALSE  |
|              |that indicates whether "suspect"      |         |        |
|              |individuals will be flagged.          |         |        |
+--------------+--------------------------------------+---------+--------+
| shrink       | A numeric value. When two consecutive|  No     |  0.10  |
|              | observations have the same trackID,  |         |        |
|              | and the ratio of size_t+1 to size_t  |         |        |
|              | is smaller than the value of         |         |        |
|              | `shrink`, the observation in year_t  |         |        |
|              | gets a `TRUE` in the "Suspect" column.|         |        |
+--------------+--------------------------------------+---------+--------+
| dormSize     | A numeric value. An individual is    |  No     | 0.05   |
|              | flagged as "suspect" if it "goes     |         |        |
|              |dormant" and has a size that is less  |         |        |
|              |than or equal to the percentile of the|         |        |
|              |size distribution for this species    |         |        |
|              |that is designated by `dormSize`      |         |        |
+--------------+--------------------------------------+---------+--------+

Throughout this vignette, we'll use a smaller subset of the `grasslandData` and
`grasslandInventory` data sets that are included in `plantTracker` for examples. 
The subset of `grasslandData` will be referred to as `dat`, because it is the 
`dat` argument in `trackSpp()`. The subset of `grasslandInventory` will be 
referred to as `inv`, since it is used for the `inv` argument.  

Here are the first few rows of the `dat` data set we'll be using:

```{r  echo = FALSE}
exampleDat <- grasslandData[grasslandData$Site == "AZ" & 
                              grasslandData$Quad == "SG2" ,]

dataShow <- head(exampleDat[, !names(exampleDat) %in%
                                  c("Clone", "Seedling", "Stems", "Basal",
                                    "sp_code_4", "Area")])
rownames(dataShow) <- NULL
(dataShow)
#knitr::kable(dataShow, caption = "**Table 1.1**: *Example `dat` data.frame*" )
```

Here are the maps for one quadrat in `dat` over the first several
years of sampling:

```{r echo = FALSE,  fig.width=7.5, fig.align = 'center', fig.cap = "**Figure 1.1**: *Spatial map of a subset of example `dat` data set*"}
exampleDat <- grasslandData[grasslandData$Site == "AZ" & 
                              grasslandData$Quad == "SG2" & 
                              grasslandData$Year %in% c(1922:1927), ]
ggplot(data = exampleDat) +
  geom_sf(aes(color = Species, fill = Species)) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 0), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 1, y = 1, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 1, xend = 1, y = 0, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  xlab("quadrat horizontal edge (m)") +
  ylab("quadrat vertical edge (m)") +
  theme_classic() +
  facet_wrap(~Year) +
  theme(axis.line = element_blank(), 
        legend.text = element_text(face = "italic"),
        plot.margin = margin(1,0,1,0)) 
```

# *2* Iterate through sites, quadrats, and species

The first step of `trackSpp()` is iterating through `dat` first by site,
then by quadrat, then by species. `inv` is also filtered down to a
single vector of sequential sampling years for the quadrat in question.
Then `trackSpp()` gets the appropriate `dorm`, `clonal`, `buff`, and
`buffGenet` arguments for that given species, either by using the
globally-specified value in the trackSpp() function call, or by extracting
the species-level value if the argument was given as a data frame of
unique values for each species. Then, the data and arguments are passed
to the `assign()` function. This function is not exported in `plantTracker`,
but the code can be accessed by typing `plantTracker:::assign()` in the
console. The remainder of this vignette describes the process of the
`assign()` function.

# *3* Track individuals over time using the `assign()` function

Once the input data has been filtered down to one site, one quadrat, and
one species, then the `assign()` function is used to track individuals
through time. In this vignette, we will use data from a site "AZs",
quadrat "SG2", and the species **"Heteropogon contortus"**. The `inv`
vector for this quadrat is
`c(1922, 1923, 1924, 1925, 1926, 1927, 1928, 1929, 1930, 1931, 1932, 1933, 1934)`

## *3.1* Get data for the first year of sampling

The data is subset yet again, this time for only the first year of
observations for this species in this quadrat, and stored in a
data frame called `tempPreviousYear`. In our example, data from 1922
will be stored in this data.frame.

## *3.2* Group genets together using `groupByGenet`, and assign "trackIDs" to each individual in the first year of sampling

Because this is the first year of sampling, no polygons have been
grouped into genets (if `clonal = TRUE`), and none have been assigned
trackIDs. Both of these tasks are accomplished by a function called
`ifClonal()`, which is internal to `assign()`. If `clonal = FALSE`, then
clonality is not allowed, and each polygon is assumed to represent a
unique genet. In this case, each polygon/row in `tempPreviousYear` is
assigned a unique "genetID" that acts as a temporary identifier that
will be used later in the function.

If `clonal = TRUE`, then clonality is allowed, and it is possible for
multiple polygons/rows in the raw data to represent one genetic
individual. In this case, we use a function called `groupByGenet()` to
group polygons together into one genet. This function uses the
`buffGenet` argument that is supplied to `trackSpp()`. The distance
(`buffGenet` x 2) is the maximum distance that two polygon edges can be
from one another and still be considered ramets from the same genet. In
other words, Any two polygons with edges that are less than (`buffGenet`
x 2) from one another will get the same "genetID." `groupByGenet()`
creates a matrix of distances between every single polygon present in
the input data.frame, and clusters them together based on proximities
that are below the threshold indicated by `buffGenet.` Then, basal area
is summed for all ramets and stored in the "basalArea_genet" column of
`tempPreviousYear`. Also, once temporary genetIDs have been assigned, a
permanent "trackID" is given to each genet. This is a combination of the
six letter species code, year of first observation, and an arbitrary
index differentiating individuals of the the same species and year of
recruitment (e.g. HETCON_1922_3). 

The following figure shows data for one year (1922) and one species 
(*Heteropogon contortus*). 

```{r echo = FALSE,  fig.show = 'hold', fig.width=7, fig.align = 'left', fig.cap = "**Figure 3.1**: *The value of 'buffGenet' used in the `trackSpp()` function can make a big difference in genetID assignments. These examples move from no genet grouping on the left, where every polygon has its own genetID, to grouping any ramets together that are less than 10 cm apart on the right. Colors and numbers indicate different genetIDs. Buffers are drawn around ramets that belong to the same genet.*", warning=FALSE}

figDat <- exampleDat[exampleDat$Species == "Heteropogon contortus" & exampleDat$Year == 1922, c("Species", "geometry")]

figDat$genetID <- as.factor(c(1:nrow(figDat)))
figDat$buffGenet <- "'clonal'=FALSE"

figDat_2 <- figDat
figDat_2$genetID <- as.factor(
  c(1,2,3,4,5,6,6,6,7,8,9,10,11,12,13, 14, 14, 15, 16, 17, 18, 19, 20, 21)
  #groupByGenet(figDat, buffGenet = .01)
  )
figDat_2$buffGenet <-  "'buffGenet'=.01"
figDat_2AG <- aggregate(x = figDat_2, 
          by = list(Species = figDat_2$Species,
            genetID = figDat_2$genetID,
            buffGenet = figDat_2$buffGenet),
          FUN = mean,
          do_union = TRUE) 
figDat_2AG <- st_buffer(figDat_2AG[,c("Species", "genetID","buffGenet","geometry")],.02)
figDat_2AG$buffGenet <- "'buffGenet'=.01"
figDat_2AG$Species <- "grouped"

figDat_3 <- figDat
figDat_3$genetID <- as.factor(
  c(1,  1,  2,  2,  2,  2,  2,  2,  3,  4,5,6,6,7,8,9,9,3,10,10,11,2,4, 4 )
  #groupByGenet(figDat, buffGenet = .05)
  )
figDat_3$buffGenet <- "'buffGenet'=.05"
figDat_3AG <- aggregate(x = figDat_3, 
          by = list(Species = figDat_3$Species,
            genetID = figDat_3$genetID,
            buffGenet = figDat_2$buffGenet),
          FUN = mean,
          do_union = TRUE)
figDat_3AG <- st_buffer(figDat_3AG[,c("Species", "genetID","buffGenet","geometry")],.05)
figDat_3AG$Species <- "grouped"
figDat_3AG$buffGenet <- "'buffGenet'=.05"
figDat <- rbind(figDat, figDat_2, figDat_3, figDat_2AG, figDat_3AG)

figDat$buffGenet <- factor(figDat$buffGenet, levels = c("'clonal'=FALSE", "'buffGenet'=.01", "'buffGenet'=.05"), ordered = TRUE)

ggplot(figDat) +
  geom_sf(data = figDat[figDat$Species == "grouped",], aes(fill = genetID), alpha = .2, color = "grey80") +
  geom_sf(data = figDat[figDat$Species == "Heteropogon contortus",], aes(fill = genetID)) + 
  #geom_sf_text(aes(label = genetID), nudge_x = .05) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 0), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 1, y = 1, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 1, xend = 1, y = 0, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  xlab("quadrat horizontal edge (m)") +
  ylab("quadrat vertical edge (m)") +
  facet_wrap(~buffGenet) +
  theme_classic() +
  scale_alpha_continuous(guide = "none") +
  scale_fill_discrete(aes(), guide = "none") +
  theme(axis.line = element_blank(), 
        plot.margin = margin(1,0,1,0)) 
```

## *3.3* Assign age and recruitment data to first year

We can also give all individuals in the first year data in the "age" and
"recruit" columns. If the first year for which there is data in `dat` is 
actually the very first year the quadrat was sampled (e.g. there are 
*Heteropogon contortus* observations in 1922, and the quadrat SG2 was first 
sampled in 1922), then we put an "NA" in both the "age" and "recruitment" 
columns. Because there was no data collected in the previous year, we don't know 
if any of these plants are new recruits, and don't know their age.

If the first year of data in `dat` -- now in `tempPreviousYear`--
is *after* the first year the quadrat was sampled (e.g. the first
*Heteropogon contortus* observations are in 1924, but the quadrat SG2
was first sampled in 1922), then we know that these individuals in
`tempPreviousYear` really are new recruits and are in their first year,
because they were not present in the previous year. They get a "1" in
both the "recruit" and "age" columns.

If the first year of data in `dat` is also the *last* year that
the quadrat is sampled (e.g. the first *Heteropogon contortus*
observations are in 1934, which is the last year of sampling), then the
observations in `tempPreviousYear` get a "1" in both the "recruit" and
"age" columns, but also get an "NA" in the "size_tplus1" and
"survives_tplus1" columns. If this is the case, the `assign()` function
still uses `ifClonal()` to assign genetIDs to these observations and then
assigns trackIDs. But there are no further steps needed to generate
demographic data, so the function returns `tempPreviousYear` as the
result after this point.

## *3.4* Compare sequential years of data to track individuals through time
Now comes the main work of the function, which compares quadrat maps for
a species over time, and assigns the same trackID to polygons that
overlap from year to year. This is accomplished using a for loop that
compares the previous year of data to the current year of data. The loop
iterates through year by the index *i*. The "previous" year is the year
with the index `i-1` in the `inv` vector, and the associated data is
stored in the `tempPreviousYear` data.frame. The "current" year is the year
with the index *i* in the `inv` vector, and the associated data is
stored in `tempCurrentYear` data.frame. There are multiple if-else
statements nested within this larger for loop, which I'll explain using
a dichotomous key below.

<a id="3_4_1"></a>

#### *3.4.1* Is there a gap between year `i-1` and year *i*?  
Not every quadrat was sampled every year, and this is indicated in the
`inv` vector. This is one case where the `dorm` argument input into
`trackSpp()` and then passed to `assign()` comes in. The value of `dorm`
indicates how many years it is "acceptable" for a plant to disappear
from the quadrat maps and still be considered the same individual with
the same trackID. The value of `dorm` must be determined by the user,
and represents a point where it's necessary to have some biological
knowledge about the species present in the data set. For example,
allowing dormancy makes sense for some species such as perennial forbs,
but doesn't for large organisms such as trees. `trackSpp()` allows you to
specify the `dorm` argument globally with one value, or individually for
each species. The `dorm` argument can also be a way to control how
"forgiving" you want to be with the data set. For example, if you expect
that plants were sometimes missed during the mapping or digitization
process, then allowing a dormancy value of "1" will help account for
this. It's important to realize that using a `dorm` value of "1" or
higher will likely *slightly* overestimate growth and survival, while
using a value of "0" will likely slightly underestimate growth and
survival.

+------------------------------------------------------------+---------+
| If a gap between `inv[i]` and `inv[i-1]` is...             |         |
+============================================================+=========+
| ... greater than the `dorm` value + 1 (e.g. if `dorm = 1`, | Go to   |
| inv[i] = 1923, and inv[i-1] = 1920; 1923 - 1920 \> (1+1)), | step    |
| then we don't know if the observations in                  |[3.4.11](#3_4_11)|
| `tempPreviousYear` survived or grew. They get an "NA" in   |         |
| the "size_tplus1" and "survives_tplus1" columns .......... |         |
+------------------------------------------------------------+---------+
| ... less than or equal to the `dorm` value + 1 (e.g. if    |         |
| `dorm = 1`, inv[i] = 1923, and inv[i-1] = 1921; 1923 -     | Proceed |
| 1921 = (1+1)), then we can compare the data from year      | to step |
| inv[i-1] (`tempPreviousYear`) to data from year inv[i]     |[3.4.2](#3_4_2)|
| (`tempCurrentYear`)  ..................................... |         |
+------------------------------------------------------------+---------+

<a id="3_4_2"></a>

#### *3.4.2* Get data for year *i*   

We already have data for the "previous" year (`inv[i-1]`) stored in
`tempPreviousYear`. Now that we know that the gap between years doesn't
exceed `dorm`, we can get data from the "current" year (`inv[i]`). We do
this by subsetting `dat` for all observations in year `inv[i]`.
Then, we use `ifClonal()` to group closely-grouped polygons into genets if
applicable, and assign genetIDs. This data set is stored in the
`tempCurrentYear` data.frame. **Proceed to step [3.4.3](#3_4_3).**

<a id="3_4_3"></a>

#### *3.4.3* Are there any observations in the "previous" year (`inv[i-1]`)?  

Even if a quadrat was sampled in `inv[i-1]`, it is possible that there
weren't actually any plants there that year.

+-----------------------------------------------------------+----------+
| If there ...                                              |          |
+===========================================================+==========+
| ... is data in `tempPreviousYear`.............            |Proceed   |
|                                                           | to step  |
|                                                           |[3.4.4](#3_4_4)|
+-----------------------------------------------------------+----------+
| ... is *not* data in `tempPreviousYear`......             | Go to    |
|                                                           | step     |
|                                                           |[3.4.12](#3_4_12)|
+-----------------------------------------------------------+----------+

<a id="3_4_4"></a>

#### *3.4.4* Add a buffer around the "previous" year data  

Now a buffer is added around each polygon in `tempPreviousYear`. This
data is stored in the `tempPreviousBuff` data.frame. This buffer is of
the width specified in the `buff` argument of `trackSpp()` that is passed
to `assign()`. Adding this buffer before comparing maps from the previous
and current years allows for mapping error and slight movement of plants
between years, which is especially likely for forbs that resprout every
year. **Proceed to step [3.4.5](#3_4_5).**

```{r echo = FALSE, warning= FALSE, fig.width=6, fig.align = 'left', fig.cap = "**Figure 3.2**: *With a 10 cm buffer, these polygons in 1922 and 1923 overlap and will be identified by trackSpp() as the **same** individual and receive the same trackID*."}

exampleDat <- grasslandData[grasslandData$Site == "AZ" & 
                              grasslandData$Quad == "SG2" & 
                              grasslandData$Year %in% c(1922,1923), ]
exampleDatIDsTemp <- exampleDat[exampleDat$Species == "Bouteloua rothrockii",]
exampleDatIDsTemp <- exampleDatIDsTemp[round(exampleDatIDsTemp$Area,7) %in% round(c( 0.0005471808, 0.0005321236),7),]
 
exampleDatIDsTemp$ghost <- "observation from current year"
exampleBuffed <- st_buffer(exampleDatIDsTemp[round(exampleDatIDsTemp$Area, 7) ==0.0005472,], dist = .10)
exampleBuffed$Year <- 1922
exampleBuffed$ghost <- "10 cm buffer"
exampleBuffedNext <- exampleBuffed
exampleBuffedNext$Year <- 1923
ghost <- exampleDatIDsTemp[round(exampleDatIDsTemp$Area, 7) == 0.0005472,]
ghost$Year <- 1923
ghost$ghost <- "polygon location in previous year"
exampleDatIDs <- rbind(exampleDatIDsTemp, ghost, exampleBuffed, exampleBuffedNext )


ggplot(data = exampleDatIDs
       ) +
  geom_sf(aes(fill = ghost, alpha = ghost, color = ghost, lty = ghost)) +
  geom_segment(aes(x = .6, xend = 1, y = .4, yend = .4), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = .6, xend = 1, y = .8, yend = .8), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = .6, xend = .6, y = .4, yend = .8), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 1, xend = 1, y = .4, yend = .8), size = .5, 
               lineend = "round", color = "grey30")+ 
  xlab("quadrat horizontal edge (m)") +
  ylab("quadrat vertical edge (m)") +
  #labs(title = Year) +
  facet_wrap( ~ Year) +
  theme_classic() +
  theme(axis.line = element_blank(), 
        legend.text = element_text(face = "italic"),
        #plot.margin = margin(1,0,1,0),
        legend.title = element_blank()) +
  scale_fill_discrete(type = c("#E69F00", "#009E73", "#A6A6A6")) +
 scale_color_discrete(type = c("#E69F00", "#009E73", "#A6A6A6")) +
  scale_alpha_discrete(range = c(0, 1, .5)) +
  scale_linetype_manual(values=c("twodash", "solid", "dotted"))
```

```{r echo = FALSE, warning= FALSE, fig.width=6, fig.align = 'left', fig.cap = "**Figure 3.3**: *With a 5 cm buffer, these polygons in 1922 and 1923 overlap and will be identified by trackSpp() as **different** individuals and receive different trackIDs.*"}
exampleDat <- grasslandData[grasslandData$Site == "AZ" & 
                              grasslandData$Quad == "SG2" , ]
exampleDatIDsTemp <- exampleDat[exampleDat$Species == "Bouteloua rothrockii" & 
                              exampleDat$Year %in% c(1922,1923),]
exampleDatIDsTemp <- exampleDatIDsTemp[round(exampleDatIDsTemp$Area,7) %in% round(c( 0.0005471808, 0.0005321236),7),]
 
exampleDatIDsTemp$ghost <- "observation from current year"
exampleBuffed <- st_buffer(exampleDatIDsTemp[round(exampleDatIDsTemp$Area, 7) ==0.0005472,], dist = .045)
exampleBuffed$Year <- 1922
exampleBuffed$ghost <- "5 cm buffer"
exampleBuffedNext <- exampleBuffed
exampleBuffedNext$Year <- 1923
ghost <- exampleDatIDsTemp[round(exampleDatIDsTemp$Area, 7) == 0.0005472,]
ghost$Year <- 1923
ghost$ghost <- "polygon location in previous year"
exampleDatIDs <- rbind(exampleDatIDsTemp, ghost, exampleBuffed, exampleBuffedNext )


ggplot(data = exampleDatIDs
       ) +
  geom_sf(aes(fill = ghost, alpha = ghost, color = ghost, lty = ghost)) +
  geom_segment(aes(x = .6, xend = 1, y = .4, yend = .4), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = .6, xend = 1, y = .8, yend = .8), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = .6, xend = .6, y = .4, yend = .8), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 1, xend = 1, y = .4, yend = .8), size = .5, 
               lineend = "round", color = "grey30")+ 
  xlab("quadrat horizontal edge (m)") +
  ylab("quadrat vertical edge (m)") +
  #labs(title = Year) +
  facet_wrap( ~ Year) +
  theme_classic() +
  theme(axis.line = element_blank(), 
        legend.text = element_text(face = "italic"),
        #plot.margin = margin(1,0,1,0),
        legend.title = element_blank()) +
  scale_fill_discrete(type = c("#E69F00", "#009E73", "#A6A6A6")) +
 scale_color_discrete(type = c("#E69F00", "#009E73", "#A6A6A6")) +
  scale_alpha_discrete(range = c(0, 1, .5)) +
  scale_linetype_manual(values=c("twodash", "solid", "dotted"))
```

<a id="3_4_5"></a>

#### *3.4.5* Are there actually any observations in the "current" year (`inv[i]`)?  

Even if a quadrat was sampled in `inv[i]`, it is possible that there
weren't actually any plants there that year. 

+----------------------------------------------+-----------------------+
| If there ...                                 |                       |
+==============================================+=======================+
| ... is data in `tempCurrentYear`.............| Proceed to step       |
|                                              | [3.4.7](#3_4_7).      |
+----------------------------------------------+-----------------------+
| ... is *not* data in `tempCurrentYear`.......| Take the entire       |
|                                              | `tempPreviousYear`    |
|                                              | data frame to step    |
|                                              | [3.4.6](#3_4_6)       |
+----------------------------------------------+-----------------------+

<a id="3_4_6"></a>

#### *3.4.6* Store observations as "ghosts" to compare to data from the next year (`inv[i+1]`) during the next iteration of the loop. 

This step also involves the "dormancy" concept discussed in section
[*3.4.1*]. If dormancy is not allowed for this species (i.e. `dorm = 0`),
then the observations in question that were "sent" to this step must be
given a "0" in the "survives_tplus1" column and an "NA" in the
"size_tplus1" column. Because they are not allowed to be dormant, if
they don't have overlapping individuals in the current year
(`inv[i]`)--which they don't if they're sent to this step--then they're
dead. **Take these observations to step [3.4.11](#3_4_11).**

However, if dormancy *is* allowed for this species, the individuals that
were "sent" to this step because they didn't overlap with anything in
year `inv[i]` can be "stored" and compared to the next set of data from
year `i+1.` We call these stored individuals "ghosts." These ghosts will
be compared to the polygons from year `i+1`, `i+2`, etc. all the way
until the dormancy argument is exceeded. For example, if some
*Heteropogon contortus* individuals were present in 1922, but did not
overlap with any plants in 1923 and `dorm = 1`, then they are stored as
"ghosts" and their locations together with those of individuals from
1923 are compared to the mapped individuals from 1924. If these "ghosts"
have no matches in the 1924 data, then they get a "0" in the
"survives_tplus1" column since they are only allowed to be dormant for
one year. We then call these individuals "dead ghosts." Any observations
that are sent to this step, but that were observed in a year that is
greater than 1 + `dorm` years ago, become "dead ghosts." The "dead
ghosts" are added to the output data.frame. **The "ghosts" are saved for the next step, which is [3.4.12](#3_4_12)**


```{r echo = FALSE, fig.width=7, fig.align = 'left', fig.cap = "**Figure 3.4**: *A visualization of the 'dormancy' scenario described above. The observation in 1922 has no overlap with any observation in 1923 (panels 1 and 2). However, if 'dorm' is greater than or equal to 1, we can save the 1922 observation as a 'ghost' (illustrated with a dotted border in panel 2). When compared to observations in 1924, there is an overlap! If 'dorm' = 1 (or more), then the observation in 1922 will get a '1' in the 'survives_tplus1' column. If 'dorm' = 0, then the observation in 1922 will get a '0' for survival, and the observation in 1924 will be a new recruit.*"}
exampleSmall <- exampleDat[exampleDat$Species == "Heteropogon contortus" & exampleDat$Year %in% c(1922, 1923, 1924),]

#117621 (1922); 117735 (1923); 117791 (1924)
exampleSmall <- exampleSmall[rownames(exampleSmall) %in% c(117621, 117735, 117791),] 
exampleSmall <- exampleSmall[rownames(exampleSmall) != 117735,]
exampleSmall <- rbind(exampleSmall, exampleSmall[1,])
exampleSmall[3,"Year"] <- 1923
exampleSmall$ghost <- c("not ghost", "not ghost", "ghost")

ggplot(data = exampleSmall) +
  geom_sf(data = exampleSmall[exampleSmall$ghost == "not ghost",] ,aes(), color = "red", fill = "red", alpha = .5) +
  geom_sf(data = exampleSmall[exampleSmall$ghost == "ghost",] ,aes(), color = "red", lty = 3, lwd = 1, fill = NA, alpha = .5) +
  geom_segment(aes(x = .5, xend = .8, y = .85, yend = .85), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = .5, xend = .8, y = 1, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = .5, xend = .5, y = .85, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = .8, xend = .8, y = 0.85, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  xlab("quadrat horizontal edge (m)") +
  ylab("quadrat vertical edge (m)") +
  xlim(c(.5,.8)) +
  ylim(c(.85,1.0)) +
  #labs(title = Year) +
  facet_wrap( ~ Year) +
  theme_classic() +
  theme(axis.line = element_blank(), 
        legend.text = element_text(face = "italic"),
        plot.margin = margin(1,0,1,0),
        legend.title = element_blank()) 
```


<a id="3_4_7"></a>

#### *3.4.7* Are there any overlaps between polygons in `tempPreviousYear` and `tempCurrentYear`?    
Use the `st_intersection` function from the sf package to determine if
there is any overlap between polygons in the the previous year
(`inv[i-1]`, stored in`tempPreviousYear`) and the current year
(`inv[i]`, stored in `tempCurrentYear`).

+-----------------------------------------------+----------------------+
| If there ...                                  |                      |
+===============================================+======================+
| ... is overlap between `tempPreviousYear` and | Proceed to step      |
| `tempCurrentYear`..................           | [3.4.8](#3_4_8)      |
+-----------------------------------------------+----------------------+
| ... is *not* overlap between                  | Take the             |
| `tempPreviousYear` and `tempCurrentYear`...   | `tempPreviousYear`   |
|                                               | data frame to step   |
|                                               | [3.4.6](#3_4_6).Take the|
|                                               | `tempCurrentYear`    |
|                                               | data frame to step   |
|                                               | [3.4.12](#3_4_12), but|
|                                               | first assign them a  |
|                                               | "1" in the "recruit" |
|                                               | column and a "1" in  |
|                                               | the "age" column.    |
+-----------------------------------------------+----------------------+


<a id="3_4_8"></a>

#### *3.4.8* Compare the overlap between `tempPreviousYear` and `tempCurrentYear` to assign trackIDs.

The `st_intersection` function used in step **[3.4.7](#3_4_7)** returns a matrix
that gives the total area of overlap between each genet in
`tempPreviousYear` and each genet in `tempCurrentYear` (the "overlap matrix"). 
There are two options from here, depending if `clonal = TRUE` or `FALSE`. 

If `clonal = TRUE`, each "parent" (those in `tempPreviousYear`) can be 
represented by more than one polygon. However, all polygons that are part of the 
same genet have the same trackID. "Child" polygons (those in `tempCurrentYear`) 
have not yet been grouped by genet, and do not have trackIDs assigned. The 
"overlap matrix" is aggregated by parent trackID so that each parent trackID has 
only one row in the matrix. The "overlap matrix" has a column for each potential 
child polygon. Each "child" polygon (those in `tempCurrentYear`) can have only 
one parent trackID (but can have multiple parent polygons). Each "parent" (those 
in `tempPreviousYear`) can have multiple child polygons. In other words, each 
row (parent) of the "overlap matrix" is allowed to have overlap values in more 
than one column, but each column (child) of the matrix can only have one overlap 
value. 

If each column of the overlap matrix has only one overlap value, then the next 
step is straightforward. Each overlapping "child" polygon is given the trackID 
of it's "parent" in the `tempCurrentYear` data frame. If there are multiple 
"children" that overlap with the same parent, those children are considered to 
be ramets of the same genet. If, however, a "child" overlaps with multiple 
parents (i.e. a column has values in more than one row), then we need to 
determine which potential "parent" is more likely the true parent. This "tie" is 
first broken by comparing the overlap area. The true "parent" is the parent with 
the highest degree of overlap with the "child". In the rare case of a tie in  
overlap area, the parent polygon with a centroid closest to the centroid of the 
child polygon is identified as the true "parent". All other values in that child 
column are turned to "NA"s. 

If `clonal = FALSE`, then each "child" can have only one "parent", and each 
"parent" only one "child". In this case, the `assign()` function uses a while 
loop to look through the matrix generated by step 3.4.7. The highest
value in the matrix indicates the greatest degree of overlap between a
given "parent" and "child." The trackID from that parent is given to
that child. Then, the overlap values in the entire "parent" row and
"child" columns in the overlap matrix are changed to zero, since each
parent can have only one child and each child can have only one parent.
The while loop repeats this process of finding the highest value in the
matrix to assign trackIDs until the entire matrix has no non-zero values
left.

Take both the `tempCurrentYear` (child) and `tempPreviousYear` (parent) data 
frames to step **[3.4.9](#3_4_9)**. 

```{r echo = FALSE, fig.width=7, fig.align = 'left', fig.cap = "**Figure 3.5**: *Here are the data for *Heteropogon contortus* in 1922 and 1923. A 5 cm buffer is shown around each genet in 1922. Data from both years have been grouped by genet using 'buffGenet' = .01*", warning = FALSE}
# comparing data from 1922 to 1923 (overlapping), then show actual trackID assignments
exampleSmall <- exampleDat[exampleDat$Species == "Heteropogon contortus" & exampleDat$Year %in% c(1922, 1923), c("Species", "Year", "geometry")]

exampleSmall$genetID <- NA

exampleSmall_22 <- exampleSmall[exampleSmall$Year == 1922,]
exampleSmall_22$genetID <- c(1,2,3,4,5,6,6,6,7,8,9,10,11,12,13,14,14,15,16,17,18,19,20,21)

exampleSmall_22 <- aggregate(x = exampleSmall_22, 
          by = list(Species = exampleSmall_22$Species,
            genetID = exampleSmall_22$genetID),
          FUN = mean,
          do_union = TRUE) 
exampleSmall_22 <- exampleSmall_22[,c("Species", "genetID", "Year", "geometry")]
exampleSmall_22$buff <- "no"
exampleSmall_22_buff <- st_buffer(exampleSmall_22, .05)
exampleSmall_22_buff$buff <- "yes"


exampleSmall_23 <- exampleSmall[exampleSmall$Year == 1923,]
exampleSmall_23$genetID <- c(1,2,3,4,5,6,7,8,9,10,11,11,12,13,14,15,16,17,18,19,20,21,22,23,24)
exampleSmall_23 <- aggregate(x = exampleSmall_23, 
          by = list(Species = exampleSmall_23$Species,
            genetID = exampleSmall_23$genetID),
          FUN = mean,
          do_union = TRUE) 
exampleSmall_23 <- exampleSmall_23[,c("Species", "genetID", "Year", "geometry")]
exampleSmall_23$buff <- "no"

exampleSmall <- rbind(exampleSmall_22, exampleSmall_22_buff, exampleSmall_23)

ggplot(data = exampleSmall) +
  geom_sf(data = exampleSmall[exampleSmall$buff == "yes",], aes(color = as.factor(Year), fill = as.factor(Year)), alpha = .2, lty = 3) +
  geom_sf(data = exampleSmall[exampleSmall$buff == "no",], aes(color = as.factor(Year), fill = as.factor(Year)), alpha = .7) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 0), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 1, y = 1, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 1, xend = 1, y = 0, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  xlab("quadrat horizontal edge (m)") +
  ylab("quadrat vertical edge (m)") +
  xlim(c(0,1)) +
  ylim(c(0,1)) +
  #labs(title = Year) +
  facet_wrap( ~ Year) +
  theme_classic() +
  theme(axis.line = element_blank(), 
        legend.text = element_text(face = "italic"),
        plot.margin = margin(1,0,1,0),
        legend.title = element_blank()) 
```

```{r echo = FALSE, fig.width=4.5, fig.align = 'left', fig.cap = "**Figure 3.6**: *Here are the buffered data for Heteropogon contortus from 1922, overlapped with the data from 1923.*", warning = FALSE}
ggplot(data = exampleSmall) +
  geom_sf(data = exampleSmall[exampleSmall$buff == "yes",], aes(color = as.factor(Year), fill = as.factor(Year)), alpha = .2, lty = 3) +
  geom_sf(data = exampleSmall[exampleSmall$buff == "no" & exampleSmall$Year == 1923,], aes(color = as.factor(Year), fill = as.factor(Year)), alpha = .7) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 0), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 1, y = 1, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 1, xend = 1, y = 0, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  xlab("quadrat horizontal edge (m)") +
  ylab("quadrat vertical edge (m)") +
  xlim(c(0,1)) +
  ylim(c(0,1)) +
  #labs(title = Year) +
  #facet_wrap( ~ Year) +
  theme_classic() +
  theme(axis.line = element_blank(), 
        legend.text = element_text(face = "italic"),
        plot.margin = margin(1,0,1,0),
        legend.title = element_blank()) 
```

```{r echo = FALSE, fig.width=7, fig.align = 'left', fig.cap = "**Figure 3.7**: *Here are the trackID assignments for these two years of data. Each trackID has a different color and a different number.*", warning = FALSE}

exampleSmall$Site <- "AZs"
exampleSmall$Quad <- "SG2"
exampleOut <- plantTracker::trackSpp(dat = exampleSmall[exampleSmall$buff == "no",], inv = list("SG2" = c(1922, 1923)), buff = .05, clonal = TRUE, dorm = 1, buffGenet = .01, aggByGenet = TRUE, printMessages = FALSE)

labels <- data.frame(trackID = unique(exampleOut$trackID), 
                     trackID_new = c(1:length(unique(exampleOut$trackID))))
exampleOut$trackID_new <- labels$trackID_new[match( exampleOut$trackID, labels$trackID)]

ggplot(data = exampleOut) +
  geom_sf(aes(color = trackID, fill = trackID), alpha = .9) +
  geom_sf_text(aes(label = trackID_new), nudge_x = .02, nudge_y = -.02) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 0), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 1, y = 1, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 1, xend = 1, y = 0, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  xlab("quadrat horizontal edge (m)") +
  ylab("quadrat vertical edge (m)") +
  xlim(c(0,1)) +
  ylim(c(0,1)) +
  scale_fill_discrete(guide = "none") +
  scale_color_discrete(guide = "none") +
  #labs(title = Year) +
  facet_wrap( ~ Year) +
  theme_classic() +
  theme(axis.line = element_blank(), 
        legend.text = element_text(face = "italic"),
        plot.margin = margin(1,0,1,0),
        legend.title = element_blank()) 
```


<a id="3_4_9"></a>

#### *3.4.9* Flag any suspect observations
If `flagSuspects = FALSE`, proceed directly to 
step **[3.4.10](#3_4_10)**. 
If `flagSuspects = TRUE`, the following checks take 
place. The first check identifies and flags any individuals in the previous year 
that became substantially smaller in the current year. For example, there are 
two overlapping observations in consecutive years that the function has given 
the same trackID. The observation in the previous year has a basal area of 20 
cm$^2$, and the observation in the current year has a basal area of 1.5 cm$^2$. 
It is possible that these two are in fact the same individual, but it is also 
possible that the observation in the current year is a new recruit that happens 
to be in a similar location to the larger plant in the previous year. If 
`flagSuspects = TRUE`, any individual from the previous year (any "parent") 
that has a basal area in the current year below a certain percentage of its size 
will be get a "TRUE" in the "Suspect" column. This threshold is defined by the 
`shrink` argument, which has a default value of 0.10 (10%). To use our previous 
example, if `shrink = .10`, the individual with a basal area of 20 cm$^2$ in the 
previous year will be flagged as "suspect" because it has shrunk to below 10% of 
its size. 

The second check flags very small individuals that go dormant. This check is 
only used if the `dorm` argument is set to "1" or higher, and if the 
observations were measured as polygons. This check can't be used for 
observations that were measured as points and converted to small polygons of a 
fixed size, since we don't know the plant's true size. A plant with a very small 
basal area is unlikely to actually survive dormancy. It is possible that the 
tracking function has correctly given the same trackID to a very small 
individual that is present in year 1, absent in year 2, and present again in 
year 3. However, it is also very possible that this very small individual died, 
and the observation in year 3 is a new recruit. This check puts a `TRUE` in the 
"suspect" column of any "parent" individual that "survives" dormancy if it is 
below a certain percentile of the size distribution for that species (which is 
created using the size data for that species provided in `dat`). The percentile 
threshold is defined by the `dormSize` argument, which has a default value 
of 0.05 (5%).

Once these checks are complete, the `tempCurrentYear` (child) and 
`tempPreviousYear` (parent) data frames go to step **[3.4.10](#3_4_10)**. 

It is important to note that, even though these checks flag individuals whose 
trackID assignment might be "suspect", the `trackSpp()` function still proceeds as 
it would if `flagSuspects` was set to `FALSE`. It is up to the user whether they 
want to exclude "suspect" observations from subsequent analyses. If you do not 
exclude these observations, it is possible that you would slightly overestimate 
survival, and underestimate recruitment and growth.

<a id="3_4_10"></a>

#### *3.4.10* Separate the "ghosts" and the new recruits from the "parent" and "child" data frames 
At this point, the `tempPreviousYear` data frame gets broken into a
`parents` data.frame, which contains data for all those genets that have
a "child" in the current year, and a `ghosts` data.frame, which contains
data for those genets that do not have a "child." The `tempCurrentYear`
data frame is broken into a `children` data.frame, containing the data
for all those genets that have a "parent" in the previous year, and an
`orphans` data.frame, which contains the data for genets that do not
have a parent.
```{r echo = FALSE, fig.width=7, fig.align = 'left', fig.cap = "**Figure 3.8**: *Here is a visualization of how the observations are broken into 'parents', 'ghosts', 'children', and 'orphans'.*", warning = FALSE}
# break d.f into parents, ghosts, children, and orphans
exampleOut$status <- NA
exampleOut[exampleOut$Year == 1922 & 
             is.na(exampleOut$survives_tplus1) == FALSE, "status"] <- "parents"
exampleOut[exampleOut$Year == 1922 & 
             is.na(exampleOut$survives_tplus1) == TRUE, "status"] <- "ghosts"
exampleOut[exampleOut$Year == 1923 & 
             is.na(exampleOut$recruit), "status"] <- "children"
exampleOut[exampleOut$Year == 1923 & 
             exampleOut$recruit == 1 & !is.na(exampleOut$recruit), "status"] <- "orphan"

ggplot(data = exampleOut) +
  geom_sf(aes(fill = status)) +
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 0), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 1, y = 1, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 1, xend = 1, y = 0, yend = 1), size = .5, 
               lineend = "round", color = "grey30") + 
  xlab("quadrat horizontal edge (m)") +
  ylab("quadrat vertical edge (m)") +
  xlim(c(0,1)) +
  ylim(c(0,1)) +
  #labs(title = Year) +
  facet_wrap( ~ Year) +
  theme_classic() +
  theme(axis.line = element_blank(), 
        legend.text = element_text(face = "italic"),
        plot.margin = margin(1,0,1,0),
        legend.title = element_blank()) 

```


**The `ghosts` data frame is sent to step [3.4.6.](#3_4_6)** All of the
observations in the `parents` data frame get a "1" in the
"survives_tplus1" column, and the total genet area of their "child" is
put into the "size_tplus1" column. Then, **the `parents` data frame is sent to step [3.4.11](#3_4_11)** All of the observations in the `children`
data frame get a "0" in the "recruit" column, the age column is
populated with 1 + the age of their parent. The observations in the
`orphans` data frame get a "1" in the "recruit" column and a "1" in the
"age" column. However if the `orphans` occur after a gap in sampling, they
instead get an "NA" in the "recruit" and "age" columns, since we don't know
whether they were recruited in year *i* or during the gap. Then, **both the**
**`children` and `orphans` data.frames are sent to step [3.4.12](#3_4_12).**


<a id="3_4_11"></a>

#### *3.4.11* Store the resulting demographic data 

Now demographic data (or NAs, if appropriate) and trackIDs have been
assigned to every individual in `tempPreviousYear` (if there are
actually observations in `inv[i-1]`), we can save these results. They
are added to a data frame that, when the for loop finishes, will be
returned by the `assign()` function. If there are any "dead ghosts", they
are also added to the output data.frame. If `inv[i]` is *not* the last
year of sampling, then proceed to step **[3.4.12](#3_4_12)**. If `inv[i]` *is* 
the last year of sampling, then the for loop is over!

<a id="3_4_12"></a>

#### *3.4.12* Get ready for the next iteration of the loop  

If there are still iterations of the loop left, that is if `inv[i]` is
not the last year of `inv`, then the data from year[i] (stored either in
`tempCurrentYear` or `children` and `orphans`) and any 'ghosts' from
previous years are put into the `tempPreviousYear` data.frame. This
happens even if `tempCurrentYear` is empty. If there are not already
genetIDs assigned to the data from `inv[i]` in `tempCurrentYear` (which
happens if this is the first year after a gap in sampling), then it is
passed through `ifClonal()`. The loop proceeds to the next value of *i*
(start again at section **[3.4.1](#3_4_1)**).

Once the loop has progressed through the 'last' year, then the output
data set will be saved, and the next species in the data set will be
sent to `assign()`!

# *4* Prepare the `trackSpp()` results to be returned!
There are just a few more steps in `trackSpp()` after the `assign()` function 
has been applied to every species present in the data set. 

## *4.1* Aggregate the results by genet
The data.frames returned by the `assign()` function are the exact same length as 
the input data frames. This means that, even though trackIDs and demographic 
data are assigned on the genet level, each 'observation', or ramet, has its own 
row of data. If the `trackSpp()` argument `aggByGenet = TRUE`, the output data 
set is passed through the `aggregateByGenet()` function from `plantTracker`. This 
aggregates the data set so that each genet in each year is represented by only 
one row of data. The polygons for each ramet are combined into one spatial 
object using the `st_union` function from the `sf` package. The resulting data 
frame will be shorter and narrower than the input data.frame, since rows are 
combined. While the output of the `assign()` function contains a column for 
"basalArea_ramet", this column is no longer present once the results 
are aggregated. 

If there were any columns in the input data frame beyond those required 
(Species, Site, Quad, Year, geometry), these will be dropped also, since the 
function can't predict whether it will be possible to aggregate those on the 
genet level. If your input data frame has data in additional columns that can be 
aggregated and that you want to keep with the demographic data, I recommend 
using `aggByGenet = FALSE`. If you want to ultimately aggregate the demographic 
data by genet, you can use the `sf::aggregate` function on your own, or modify 
the code for the `aggregateByGenet()` function to include your additional columns. 
If you have set `clonal = FALSE` for all species in your input data.frame, I 
also recommend using `aggByGenet = FALSE`, since your results will already be on 
the genet scale! 

## *4.2* Informative messages
If the argument `printMessages = TRUE`, one or two messages will be printed as 
each species goes through the `assign()` function. These messages are not 
warnings or errors! Unless the function returns a message preceded by the word 
"warning" or "error", the function is working! The messages I'm talking about 
here provide information about why there are "NA"s present in the demographic 
results, which may be concerning if you aren't expecting them. The first message 
tells you which year is the last year of sampling for this quadrat. Observations 
in the last year of sampling will have an "NA" in both the "survives_tplus1" and 
"size_tplus" columns because we have no data to determine whether they survived. 
The second message only appears if there is a gap in sampling for that quadrat 
that exceeds the `dorm` argument. The message indicates that observations in the 
year(s) preceding that gap will have "NA"s in the "survives_tplus1" 
"size_tplus1" columns, since we don't know when they died. If both 
`printMessages = TRUE` and `aggByGenet = TRUE`, an additional message will be 
printed. This message will warn that the output data frame is shorter and 
narrower than the input data.frame, and will explain why that is. 

Lastly, if `printMessages = TRUE`, the `trackSpp()` function will print progress 
messages that indicate which site is being run the function, then which species, 
then which quadrat. This is helpful both to know how far the function has gotten 
in your data, and also is helpful if the function errors out. You can find 
roughly where the problem in the data is, since you know the species, quadrat, 
and site where the function crashed. 

If `printMessages = FALSE`, then no messages will be returned. 

# *5* Examples
Here are the trackID assignments for 4 years of observations of 
*Heteropogon contortus* from a subset of the SG2 quadrat near Tucson, Arizona. 
The `trackSpp` function here uses `dorm = 1`, `clonal = TRUE`, `buff = 0.05` and 
`buffGenet = 0.01`.  
```{r echo = FALSE, fig.width=7, fig.align = 'left', fig.cap = "**Figure 5.1**: *Here are the trackID assignments over 4 years of data. Each trackID is shown as a different color and has a different number.*", warning = FALSE}
exampleSmall <- exampleDat[exampleDat$Species == "Heteropogon contortus" &
                           exampleDat$Year %in% c(1922:1925),]
exampleSmall$Site <- "AZs"
exampleSmall$Quad <- "SG2"

## trim the dataset to be smaller
## make a bounding box
pl = list(rbind(c(0,0), c(.5,0), c(.5,.7), c(0,.7), c(0,0)))
box <- st_polygon(pl)
exampleSmall <-suppressWarnings(st_intersection(exampleSmall, box))

## get trackIDs 
exampleOut <- suppressMessages(plantTracker::trackSpp(dat = exampleSmall, inv = list("SG2" = c(1922:1925)), buff = .05, clonal = TRUE, dorm = 1, buffGenet = .01, aggByGenet = TRUE, printMessages = FALSE))

labels <- data.frame(trackID = unique(exampleOut$trackID), 
                     trackID_new = c(1:length(unique(exampleOut$trackID))))
exampleOut$trackID_new <- labels$trackID_new[match( exampleOut$trackID, labels$trackID)]

ggplot(data = exampleOut) +
  geom_sf(aes(color = trackID, fill = trackID), alpha = .9) +
  geom_sf_text(aes(label = trackID_new), nudge_x = .04) +
  geom_segment(aes(x = 0, xend = .5, y = 0, yend = 0), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = .5, y = .7, yend = .7), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = .7), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = .5, xend = .5, y = 0, yend = .7), size = .5, 
               lineend = "round", color = "grey30") + 
  xlab("quadrat horizontal edge (m)") +
  ylab("quadrat vertical edge (m)") +
  xlim(c(0,.5)) +
  ylim(c(0,.7)) +
  scale_fill_discrete(guide = "none") +
  scale_color_discrete(guide = "none") +
  #labs(title = Year) +
  facet_wrap(~ Year, ncol = 4) +
  theme_classic() +
  theme(axis.line = element_blank(), 
        legend.text = element_text(face = "italic"),
        plot.margin = margin(1,0,1,0),
        legend.title = element_blank()) 
```

Here is that same data subset, but run through `trackSpp()` using a buffGenet 
value of .05
```{r echo = FALSE, fig.width=7, fig.align = 'left', fig.cap = "**Figure 5.2**: *Here are the trackID assignments over 4 years of data. Each trackID is shown as a different color and has a different number.*", warning = FALSE}
## get trackIDs 
exampleOut <- suppressMessages(plantTracker::trackSpp(dat = exampleSmall, inv = list("SG2" = c(1922:1925)), buff = .05, clonal = TRUE, dorm = 1, buffGenet = .05, aggByGenet = TRUE, printMessages = FALSE))

labels <- data.frame(trackID = unique(exampleOut$trackID), 
                     trackID_new = c(1:length(unique(exampleOut$trackID))))
exampleOut$trackID_new <- labels$trackID_new[match( exampleOut$trackID, labels$trackID)]

ggplot(data = exampleOut) +
  geom_sf(aes(color = trackID, fill = trackID), alpha = .9) +
  geom_sf_text(aes(label = trackID_new), nudge_x = .04) +
  geom_segment(aes(x = 0, xend = .5, y = 0, yend = 0), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = .5, y = .7, yend = .7), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = .7), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = .5, xend = .5, y = 0, yend = .7), size = .5, 
               lineend = "round", color = "grey30") + 
  xlab("quadrat horizontal edge (m)") +
  ylab("quadrat vertical edge (m)") +
  xlim(c(0,.5)) +
  ylim(c(0,.7)) +
  scale_fill_discrete(guide = "none") +
  scale_color_discrete(guide = "none") +
  #labs(title = Year) +
  facet_wrap(~ Year, ncol = 4) +
  theme_classic() +
  theme(axis.line = element_blank(), 
        legend.text = element_text(face = "italic"),
        plot.margin = margin(1,0,1,0),
        legend.title = element_blank()) 
```


Here it is again, but run through `trackSpp()` using `clonal = FALSE`
```{r echo = FALSE, fig.width=7, fig.align = 'left', fig.cap = "**Figure 5.3**: *Here are the trackID assignments over 4 years of data. Each trackID is shown as a different color and has a different number.*", warning = FALSE}

## get trackIDs 
exampleOut <- suppressMessages(plantTracker::trackSpp(dat = exampleSmall, inv = list("SG2" = c(1922:1925)), buff = .05, clonal = FALSE, dorm = 1, aggByGenet = TRUE, printMessages = FALSE))

labels <- data.frame(trackID = unique(exampleOut$trackID), 
                     trackID_new = c(1:length(unique(exampleOut$trackID))))
exampleOut$trackID_new <- labels$trackID_new[match( exampleOut$trackID, labels$trackID)]

ggplot(data = exampleOut) +
  geom_sf(aes(color = trackID, fill = trackID), alpha = .9) +
  geom_sf_text(aes(label = trackID_new), nudge_x = .04) +
  geom_segment(aes(x = 0, xend = .5, y = 0, yend = 0), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = .5, y = .7, yend = .7), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = .7), size = .5, 
               lineend = "round", color = "grey30") + 
  geom_segment(aes(x = .5, xend = .5, y = 0, yend = .7), size = .5, 
               lineend = "round", color = "grey30") + 
  xlab("quadrat horizontal edge (m)") +
  ylab("quadrat vertical edge (m)") +
  xlim(c(0,.5)) +
  ylim(c(0,.7)) +
  scale_fill_discrete(guide = "none") +
  scale_color_discrete(guide = "none") +
  #labs(title = Year) +
  facet_wrap(~ Year, ncol = 4) +
  theme_classic() +
  theme(axis.line = element_blank(), 
        legend.text = element_text(face = "italic"),
        plot.margin = margin(1,0,1,0),
        legend.title = element_blank()) 
```
